The Evolution of Phenotypically Invariant Genetic Networks

### 2017-11-20 kalman_walk_paper.tex
discussion edits

% it says something about speciation and DMIs
We have suggested an interpretation of system nonidentifiability: to see it as an evolutionarily neutral manifold, and not simply a computational nuisance. 
We have demonstrated a method to analytically determine the set of all phenotypically equivalent gene networks; 
by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition.
%Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection. 
This set is explored over evolutionary time when phenotype is conserved, and can lead to a diverse set of consequences, 
including an increase in hybrid inviability.
We emphasize that here, hybrid inviability is a consequence of recombining different, yet functionally equivalent, mechanisms.
\plr{See refs in Barton 2010 paper.}

@@

% summary and consequences of main result on speed of speciation
After demonstrating that it is, in principle, possible for phenotypically indistinguishable populations to speciate, we next probed whether system drift is possible under neutrality and whether the processs is fast enough to be a significant driver of speciation. This analysis was carried out by applying methods from quantitative genetics. In this framework a population is composed of many organisms, some with slightly differing gene regulatory network architectures. Some of this intrapopulation architectural variation is neutral and some is mildly deleterious and contributes to genetic load. Since there are always numerous neighboring system architecures with equivalent phenotypes and populations are not genetically homogenous, the typical system architecture within a population will drift. The speed at which a population will drift per generation, and thus the rate at which hybrid inviability appears, is a function of a population's size and of how much intrapopulation variation is present. Above, with the input of a few population parameters, we predict the rate at which isolated populations' hybrids become inviable. Our model suggests, that for a biologically reasonable range of parameter values, system drift can be a significant, if not rapid, driver of speciation, under neutrality.

Additionally we see $F_2$ hybrid fitnesses plummet much faster than that of $F_1$ hybrids. 
%\plr{refer to Turelli here}
This is due to recombination shuffling system architectures leaving the hybrid system homozygous for one parent at one particular locus and homozygous for the other parent at another incompatible locus. If part of a gene regulatory network resides on the X chromosome, then $F_1$ males (XY) will effectivley look like $F_2$s at those locations. This observation is consistent with Haldane's rule; that if only one hybrid sex is inviable or sterile it is likely to be the heterogametic sex.

Lastly, we show that hybrid gene networks, under neutral processes, break down as a function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding, hedges2015tree}.

While sumarizing speciation research \citet{orr2004speciation} remarked that ``speciation results from positive Darwinian selection within species'' thus maintaining one of the ``central tenets of the modern synthesis''. While it is known that divergent selection and genetic conflict, as well as other mechanisms, can precipitate speciation, we have shown here that, surprisingly, neutral drfit may also significantly drive species formation.
%, although it is at present still unclear the relative contribution of each process. 
System drift induced hybrid breakdown occurs rapidly even in the absence of postive Darwinian selection, suggesting that species originate not necessarily by means of natural selection \citep{darwin1859origin}. While the incompatibility of any one given network within an organism may be small, we note that organsisms are made up of many different networks, and that the cumulative impact of system drift across all of these may be substantial. Further, even in populations experiencing directional selection, networks underlying conserved parts of the phenotype can still experience neutral drift and harbor incompatibilities.

Despite the results of this inquiry, in practice it remains unclear whether system drift or other speciation mechanisms predominate, as hybrid incompatibilities due to both selection and conflict have been uncovered and hint at general processes \citep{orr2004speciation}. It also seems reasonable that system drift could be common as there is evidence (1) that the set of molecular mechanisms underlying a trait is degenerate (illustrated in this paper), (2) that there is significant regulatory variation segregating within populations, (3) that functionally equivalent systems, even in closesly related species, can be mechanistically divergent \citep{dalal2017transcription}, (4) transcription in hybrids between closely related species with conserved transcriptional patterns can be divergent \citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}, that (5) gene misregulation can cause dysfunction (by definition), and (6) that numerous cryptic species complexes exist (e.g. sun skinks \citep{barley2013challenge}).

Additionally, if we ignore the fitness function used in this paper, our model still predicts that $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time with parentals, respectivey. This has implications for experimentation, as one could plot the change in phenotype (or gene expression) as a function of divergence, to test the predictions of this model.

>>>>

% it says something about speciation and DMIs
We have suggested an interpretation of system nonidentifiability: to see it as an evolutionarily neutral manifold, and not simply a computational nuisance. 
We have demonstrated a method to analytically determine the set of all phenotypically equivalent gene networks; 
by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition.
%Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection. 
%% summary and consequences of main result on speed of speciation
%After demonstrating that it is, in principle, possible for phenotypically indistinguishable populations to speciate, we next probed whether system drift is possible under neutrality and whether the processs is fast enough to be a significant driver of speciation. This analysis was carried out by applying methods from quantitative genetics. In this framework a population is composed of many organisms, some with slightly differing gene regulatory network architectures. Some of this intrapopulation architectural variation is neutral and some is mildly deleterious and contributes to genetic load. Since there are always numerous neighboring system architecures with equivalent phenotypes and populations are not genetically homogenous, the typical system architecture within a population will drift. 
This set is explored by a population over evolutionary time while phenotype is conserved, and can lead to speciation via an increase in hybrid inviability.
The speed at which a population will drift through this set, and thus accrue hybrid incompatibilities, is a function of effective population size and genetic variation. 
With the input of a range of biologically reasonable parameter values, our model predicts that system drift will be a significant -- and likely a rapid -- driver of speciation.
We emphasize that here, hybrid inviability is a consequence of recombining different, yet functionally equivalent, mechanisms. 
\plr{See refs in Barton 2010 paper.}

Consistent with empirical observation, we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s (as $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectivey).
%\plr{refer to Turelli here}
In this model, this is due to recombination shuffling system architectures, leaving the $F_2$ system homozygous for one parent at one particular locus and homozygous for the other parent at another possibly incompatible locus.
Interestingly, if part of a gene regulatory network resides on the sex chromosomes (X), then heterogametic $F_1$s (typically males) will effectivley look like $F_2$s. This observation is consistent with, and may in fact explain, Haldane's rule: that if only one hybrid sex is inviable or sterile it is likely to be the heterogametic sex.

@@

Our results show that hybrids, under neutral processes, break down as a function of genetic distance, and despite the present model's myriad simplifications and assumptions, 
it may, in part, explain the observed broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding, hedges2015tree}. \js{maybe also cite the phylogentics lit here? Like the speciation clock stuff.}

Although, as classically formulated, the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary}, \citet{orr2004speciation} remarks that hybrid incompatibility data suggests that ``speciation results from positive Darwinian selection within species'' thus maintaining one of the ``central tenets of the modern synthesis''. 
%In practice it remains unclear whether system drift or other speciation mechanisms predominate, 
%as hybrid incompatibilities due to both selection and conflict have been uncovered and hint at general processes \citep{orr2004speciation}. 
While it is known that divergent selection and genetic conflict, as well as other mechanisms, can precipitate speciation, we have shown here that, surprisingly, neutral drfit may also significantly drive species formation.
This seems reasonable as there is evidence that the set of molecular mechanisms underlying a trait is degenerate, that there is significant regulatory variation segregating within populations, that functionally equivalent systems, even in closesly related species, can be mechanistically divergent \citep{dalal2017transcription}, transcription in hybrids between closely related species with conserved transcriptional patterns can be divergent \citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}, that gene misregulation can cause dysfunction (by definition), and that numerous cryptic species complexes exist (e.g. sun skinks \citep{barley2013challenge}).

If system drift induced hybrid breakdown occurs as rapidly as our model sugests, even in the absence of postive Darwinian selection, this would imply that species originate not necessarily by means of natural selection \citep{darwin1859origin}. While the incompatibility of any one given network within an organism may be small, we note that organsisms are made up of many different networks, and that the cumulative impact of system drift across all of these may be substantial. Further, even in populations experiencing directional selection, networks underlying conserved parts of the phenotype can still experience neutral drift and harbor incompatibilities.
